Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-08 paper.tex
mkl going through paper

\title{Cooling Requirements for the Vertical Shear Instability 
  in Protoplanetary Disks}
%\title{Vertical shear instability in astrophysical disks: vertically
%global linear calculations with thermal relaxation and applications
%to the Solar Nebula}  
\author{Min-Kai Lin\altaffilmark{$\dagger$} \& Andrew N Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\altaffiltext{$\dagger$}{Steward Theory Fellow}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

\begin{abstract}
  It is difficult to understand how cold circumstellar disks accrete onto their central stars.  Due to 
  lower levels of ionization, magnetic transport mechanisms that
  operate in warmer disks are less effective.  
  A hydrodynamic mechanism, the vertical shear instability (VSI), offers a means to drive angular momentum 
  transport in cold accretion disks such as protoplanetary disks (PPDs).  
  The VSI is driven by a weak vertical gradient in the disk's orbital motion.
  In order to grow, the VSI must overcome vertical buoyancy, a strong
  stabilizing influence in cold disks, where  
  heating is dominated by external irradiation.  Rapid cooling, via
  radiative losses, reduces the effective buoyancy 
  and allows the VSI to operate.  In this paper, we quantify the
  cooling timescale, $t_c$, needed for growth of the VSI.
   We perform a linear analysis of the VSI with cooling in
  vertically global and radially local disk models. 
  For irradiated disks, we find that the VSI is most vigorous for
  rapid cooling with $t_c < \OmK^{-1}  h |q| / (\gamma -1)$ 
  in terms of the Keplerian orbital frequency, $\OmK$; the disk's
  aspect ratio, $ h \ll 1$; the radial  
  powerlaw temperature gradient, $q$; and the adiabatic index,
  $\gamma$.  For longer cooling times, the VSI  
  is much less effective because growth slows and shifts to smaller length scales, which are more prone to 
  viscous or turbulent decay.  We apply our results to PPD models 
  where $t_c$ is determined by the opacity of dust grains.  We
  find that the VSI is most effective at intermediate radii, from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.
  Growth is suppressed by long cooling times both in the opaque  
  inner disk and the optically thin outer disk.  A reduction in the
  dust opacity by a factor of 10 increases cooling times  
  enough to quench the VSI at all disk radii.  Thus the formation of
  solid protoplanets, a sink for dust grains, can impede the VSI. 
\end{abstract}

@@

This paper is organized as follows. In \S\ref{vsi_require} we explain, without derivation,  
our main result for the critical cooling timescale, beyond which VSI growth is 
suppressed. %We then present the  
%formal linear problem in \S\ref{setup}---\ref{numerical}.    
 We develop our disk model in \S\ref{setup} and derive linearized
 perturbation equations in  \S\ref{linear}.
 Section \ref{analytical} contains our main analytic results, leading to 
 the derivation of the critical cooling time in \S\ref{iso_vsi_beta_crit}.
 In \S\ref{numerical} we analyze linear VSI growth, and numerically confirm the critical cooling time.
We apply our results to PPDs in \S\ref{application}, with cooling times derived from dust opacities.
We discuss caveats and extensions in \S\ref{caveats}.
 We conlude in \S\ref{summary} with a summary.  Some technical and background 
 developments are explained in the appendices.

@@

We are more interested, however, the the global instability criterion for disturbances at all heights.
Our desired result 
is the cooling requirement 
\begin{align}\label{prelim_bcrit}
  t_c <  &\underbrace{\frac{h |q| }{\gamma - 1}}_{\beta_\mathrm{crit}}\OmK^{-1}
\end{align}   
for VSI growth that we derive in \S\ref{iso_vsi_beta_crit}.  We can simply 
(but non-rigourously) obtain Eq. \ref{prelim_bcrit}
  by evaulating  Eq. \ref{bcrit_est} at $|z|=\gamma H/2$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.7cm 0cm
  0.73cm]{figures/bcrit_mmsn2d_kx10}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.46cm 0cm
  0.73cm]{figures/bcrit_mmsn2d_kx50}
  \caption{Estimate of the thermal timescale  in a fiducial 
    PPD model, the Minimum Mass Solar Nebula. The red
    lines corresponds to a maximum thermal timescale, derived in
    this work, below which the VSI is expected to operate effectively
    (the region bounded by the red lines).     
    \label{bcrit_mmsn2d} 
  }
\end{figure}

We now consider whether, and where, thermal timescales in realistic PPDs
are short enough to support VSI growth.  Fig. \ref{bcrit_mmsn2d} plots the
 thermal timescales in our fiducial PPD model, see \S\ref{toy_relax} for details.
 The regions between the red lines satisfy the cooling requirement of Eq.\ \ref{prelim_bcrit}.
 Even without a detailed analysis we can expect VSI growth between $\sim 10$ -- 150 AU.
 The inner disk is complicated by the fact that the optically thick midplane has longer cooling times. 
 Thus the cooling criterion fails in the midplane but is satisfied above.  This case requires the 
 more detailed analysis of \S\ref{application}.

  Fig. \ref{bcrit_mmsn2d} also compares two different radial wavenumbers, 
  $k_x = 10/H$ and $50/H$.  In the inner disk, the higher wavenumber mode cools faster, due to radiative diffusion.
  Thus in the inner disk, shorter wavelength VSI modes are more like to grow. In the outer disk, the cooling time is the 
  same for different $k_x$ and $z$, because optically thin cooling is independent of both lengthscale and density.
  Thus the outer barrier to growth, beyond $\sim 150$ AU in this model, is independent of  wavelength. 
This example highlights the fact that optically thin cooling sets a lower limit to the cooling time; a fact that is sometimes obscured when 
the radiative diffusion approximation is made at the outset.

>>>>

\title{Cooling Requirements for the Vertical Shear Instability 
  in Protoplanetary Disks}
%\title{Vertical shear instability in astrophysical disks: vertically
%global linear calculations with thermal relaxation and applications
%to the Solar Nebula}  
\author{Min-Kai Lin\altaffilmark{1} \& Andrew N Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\altaffiltext{1}{Steward Theory Fellow}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

\begin{abstract}
  It is difficult to understand how cold circumstellar disks accrete onto their central stars.  Due to 
  lower levels of ionization, magnetic transport mechanisms that
  operate in warmer disks are less effective.  
  A hydrodynamic mechanism, the vertical shear instability (VSI), offers a means to drive angular momentum 
  transport in cold accretion disks such as protoplanetary disks (PPDs).  
  The VSI is driven by a weak vertical gradient in the disk's orbital motion.
  In order to grow, the VSI must overcome vertical buoyancy, a strongly
  stabilizing influence in cold disks, where  
  heating is dominated by external irradiation.  Rapid cooling, via
  radiative losses, reduces the effective buoyancy 
  and allows the VSI to operate.  In this paper, we quantify the
  cooling timescale, $t_c$, needed for growth of the VSI.
   We perform a linear analysis of the VSI with cooling in
  vertically global and radially local disk models. 
  For irradiated disks, we find that the VSI is most vigorous for
  rapid cooling with $t_c < \OmK^{-1}  h |q| / (\gamma -1)$ 
  in terms of the Keplerian orbital frequency, $\OmK$; the disk's
  aspect ratio, $ h \ll 1$; the radial  
  power-law temperature gradient, $q$; and the adiabatic index,
  $\gamma$.  For longer cooling times, the VSI  
  is much less effective because growth slows and shifts to smaller length scales, which are more prone to 
  viscous or turbulent decay.  We apply our results to PPD models 
  where $t_c$ is determined by the opacity of dust grains.  We
  find that the VSI is most effective at intermediate radii, from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.
  Growth is suppressed by long cooling times both in the opaque  
  inner disk and the optically thin outer disk.  A reduction in the
  dust opacity by a factor of 10 increases cooling times  
  enough to quench the VSI at all disk radii.  Thus the formation of
  solid protoplanets, a sink for dust grains, can impede the VSI. 
\end{abstract}

@@

This paper is organized as follows. In \S\ref{vsi_require} we explain, without derivation,  
our main result for the critical cooling timescale, beyond which VSI growth is 
suppressed. %We then present the  
%formal linear problem in \S\ref{setup}---\ref{numerical}.    
 We develop our disk model in \S\ref{setup} and derive linearized
 perturbation equations in  \S\ref{linear}.
 Section \ref{analytical} contains our main analytic results, leading to 
 the derivation of the critical cooling time in \S\ref{iso_vsi_beta_crit}.
 In \S\ref{numerical} we analyze linear VSI growth, and numerically confirm the critical cooling time.
We apply our results to PPDs in \S\ref{application}, with cooling times derived from dust opacities.
We discuss caveats and extensions in \S\ref{caveats}.
 We conclude in \S\ref{summary} with a summary.  Some technical and background 
 developments are explained in the appendices.

@@

We are more interested, however, in the global instability criterion for disturbances at all heights.
Our key result 
is the cooling requirement 
\begin{align}\label{prelim_bcrit}
  t_c <  &\frac{h |q| }{\gamma - 1}\OmK^{-1}
\end{align}   
for VSI growth that we derive in \S\ref{iso_vsi_beta_crit}.  We can simply 
(but non-rigorously) obtain Eq. \ref{prelim_bcrit}
by evaluating  Eq. \ref{bcrit_est} at $|z|=\gamma H/2$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.7cm 0cm
  0.73cm]{figures/bcrit_mmsn2d_kx10}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.46cm 0cm
  0.73cm]{figures/bcrit_mmsn2d_kx50}
  \caption{Estimate of the thermal timescale $t_c$  in a fiducial 
    PPD model, the Minimum Mass Solar Nebula. Here, $\beta_\mathrm{crit}=h|q|/(\gamma-1)$  
    is a maximum dimensionless cooling timescale below which the VSI
    is expected to operate effectively. This corresponds to   
    the region bounded by the red lines. 
    \label{bcrit_mmsn2d} 
  }
\end{figure}

We now consider whether, and where, thermal timescales in realistic PPDs
are short enough to support VSI growth.  Fig. \ref{bcrit_mmsn2d} plots the
 thermal timescales in our fiducial PPD model, see \S\ref{toy_relax} for details.
 The regions between the red lines satisfy the cooling requirement of Eq.\ \ref{prelim_bcrit}.
 Even without a detailed analysis we can expect VSI growth between $\sim 10$ -- 150AU.
 The inner disk is complicated by the fact that the optically thick
 midplane has longer cooling times.  
 Thus the cooling criterion fails in the midplane but is satisfied
 above.  This case requires the  
 more detailed analysis of \S\ref{application}.

  Fig. \ref{bcrit_mmsn2d} also compares two different radial wavenumbers, 
  $k_x = 10/H$ and $50/H$.  In the inner disk, the higher wavenumber
  mode cools faster, due to radiative diffusion. 
  Thus in the inner disk, shorter wavelength VSI modes are more likely
  to grow. In the outer disk, the cooling time is the  
  same for different $k_x$ and $z$, because optically thin cooling is
  independent of both lengthscale and density. 
  Thus the outer barrier to growth, beyond $\sim 150$ AU in this
  model, is independent of  wavelength.  
This example highlights the fact that optically thin cooling sets a
lower limit to the cooling time; a fact that is sometimes obscured
when  
the radiative diffusion approximation is made at the outset.
